---
title: "Quantify, correct, and design"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quantify, correct, and design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4)
library(spLeakage)
```

## From "is it leaking?" to "what do I do about it?"

Detecting leakage is only the first step. Once you know a split leaks, four
practical questions follow, and `spLeakage` answers each:

1. **How inflated is my reported accuracy?** → `estimate_optimism()` (refit) and
   `predict_optimism()` (fast emulator).
2. **What number should I report instead?** → `deleak_estimate()` — a corrected
   accuracy with a confidence interval, even for a *published* result.
3. **Which validation scheme should I have used?** → `rank_cv_schemes()`.
4. **Where should I collect independent validation data?** → `design_validation()`.

We reuse the clustered example throughout.

```{r data}
set.seed(1)
nc <- 8; ce <- cbind(runif(nc, .1, .9), runif(nc, .1, .9)); cl <- sample(nc, 400, TRUE)
xy <- ce[cl, ] + cbind(rnorm(400, 0, .04), rnorm(400, 0, .04))
d <- data.frame(x = pmin(pmax(xy[, 1], 0), 1), y = pmin(pmax(xy[, 2], 0), 1))
d$z <- sin(2 * pi * d$x) + cos(2 * pi * d$y) + rnorm(400, 0, .15)
folds <- sample(rep_len(1:10, 400))
grid <- as.matrix(expand.grid(x = seq(0, 1, .07), y = seq(0, 1, .07)))
tgt <- prediction_target(grid = grid, type = "grid")
```

## 1. Quantify the optimism (the refit route)

`estimate_optimism()` re-evaluates the model under a leakage-controlled,
design-matched scheme and reports the gap to your reported (random-CV) error.

```{r}
estimate_optimism(d, folds, response = "z", coords = c("x", "y"))
```

The `optimism` line is the headline: it gives the absolute gap *and* the percentage
inflation of your accuracy. A large positive percentage means the random split made
the model look that much better than the design-matched scheme says it is. (For a
true probability sample this number can be negative — pessimism — and is reported as
such.)

### Proper scoring rules for non-Gaussian responses

For prevalence (binary) or count responses, RMSE is not the right metric. Pass a
proper scoring rule so the optimism is meaningful — `brier`/`logloss` for binary,
`poisson` for counts:

```{r}
db <- d; db$z <- rbinom(400, 1, plogis(2 * sin(2 * pi * d$x)))
estimate_optimism(db, folds, response = "z", coords = c("x", "y"), metric = "brier")
```

## 2. Quantify the optimism (the fast emulator route)

The refit route needs to re-run the model. `predict_optimism()` instead uses an
emulator **pre-calibrated on a large simulation study** and shipped with the
package, returning an estimate (with a 90% interval) in milliseconds and *without*
refitting.

```{r}
predict_optimism(d, folds, tgt, response = "z", coords = c("x", "y"), model = "idw")
```

Two things to note. First, when the query is in range the emulator returns a point
estimate with a genuine (conformal) predictive interval, not a point guess. Second —
and this is deliberate — the emulator carries its **own** area-of-applicability
guard. In fact this small, tightly clustered example falls *outside* the envelope the
emulator was trained on, so it prints `optimism: NA [outside AOA: DI ... > ...]` and
**refuses to answer** rather than extrapolate. A leakage tool must not commit the sin
it polices. When the emulator abstains like this, fall back to the refit route in
section 1 (`estimate_optimism()`), which always applies.

## 3. Correct it: the de-leaked estimate

`deleak_estimate()` is the number you should actually report: a bias-corrected
("de-leaked") accuracy with a bootstrap confidence interval. Pass your reported
metric value via `reported=` and it will also translate *that specific number* into
its corrected counterpart — so you can correct a published result without the
original model or refitting.

```{r}
deleak_estimate(d, folds, response = "z", coords = c("x", "y"),
                reported = 0.30, n_boot = 200)
```

* `de-leaked estimate` (with CI) is the corrected accuracy.
* The `inflated by ...%` line restates the optimism as a multiplier on your error.
* The final line maps your supplied `reported` 0.30 to its de-leaked value.

For extrapolation deployments, pass `target=` so the correction uses a
**target-matched** control (the paper shows this recovers ~89% of the true
independent error, versus ~58% with a fixed block control).

## 4. Recommend and rank validation schemes

`recommend_validation()` gives design-aware guidance (recall: the design is
*declared*, never inferred):

```{r}
recommend_validation(estimand = "prediction", design = "clustered", target = "grid")
```

Where `recommend_validation()` gives expert guidance, `rank_cv_schemes()` makes it
**data-driven**: it builds each candidate scheme on *your* data and ranks them by how
closely each one's geometry matches your declared deployment target (smallest
`|SLI_rho|` wins).

```{r}
rank_cv_schemes(d, tgt, response = "z", coords = c("x", "y"))
```

The `*` marks the recommended scheme. On clustered data you will typically see
random CV flagged optimistic, buffered CV *over-corrected into pessimism*, and block
CV closest to deployment — a nuance a static lookup table cannot provide.

## 5. Design where to collect validation data

The honest way to know your true map accuracy is to measure it on **independent**
points. Given a budget, where should those points go? `design_validation()` inverts
the leakage logic: it stratifies space by distance-to-training (on the deployment
distribution), models the per-stratum error from the spatial theory, and allocates
the budget by Neyman allocation with inclusion weights, so the resulting
weighted-mean accuracy estimate is both unbiased and minimum-variance.

```{r}
design_validation(d, tgt, budget = 25, response = "z", coords = c("x", "y"))
```

The `SE ... lower than simple random validation` line quantifies the saving: in the
paper's experiments the optimal design reaches the same precision from a budget of
30 points that random placement needs far more to match. The returned strata table
(deployment weight × mean error × points placed) shows where the budget went and the
inclusion weights to use when averaging the validation errors.

## Recap

`spLeakage` takes you from a single leakage score to a defensible, corrected
accuracy estimate and a plan for collecting the data that would confirm it. Combined
with the multi-channel detectors in `vignette("channels")` and the conceptual
framing in `vignette("spLeakage")`, that is the full diagnostic workflow.
```
