---
title: "Quantify, correct, and design"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quantify, correct, and design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(spLeakage)
```

Detecting leakage is the first step. `spLeakage` also quantifies the *consequence*
(how inflated your accuracy is), corrects it, recommends a better scheme, and designs
where to collect independent validation data.

```{r data}
set.seed(1)
nc <- 8; ce <- cbind(runif(nc, .1, .9), runif(nc, .1, .9)); cl <- sample(nc, 400, TRUE)
xy <- ce[cl, ] + cbind(rnorm(400, 0, .04), rnorm(400, 0, .04))
d <- data.frame(x = pmin(pmax(xy[, 1], 0), 1), y = pmin(pmax(xy[, 2], 0), 1))
d$z <- sin(2 * pi * d$x) + cos(2 * pi * d$y) + rnorm(400, 0, .15)
folds <- sample(rep_len(1:10, 400))
grid <- as.matrix(expand.grid(x = seq(0, 1, .07), y = seq(0, 1, .07)))
tgt <- prediction_target(grid = grid, type = "grid")
```

## Quantify the optimism

How much is the reported accuracy inflated, relative to a leakage-controlled scheme?

```{r}
estimate_optimism(d, folds, response = "z", coords = c("x", "y"))
```

For binary (disease-prevalence) or count responses, use a proper scoring rule:

```{r}
db <- d; db$z <- rbinom(400, 1, plogis(2 * sin(2 * pi * d$x)))
estimate_optimism(db, folds, response = "z", coords = c("x", "y"), metric = "brier")
```

## Correct it: the de-leaked estimate

The de-leaked estimate is the accuracy you should report, with a confidence interval.
It can even correct a *published* number without refitting.

```{r}
deleak_estimate(d, folds, response = "z", coords = c("x", "y"),
                reported = 0.30, n_boot = 200)
```

## Fast emulator route

`predict_optimism()` uses a calibrated emulator (no refitting); it refuses queries
outside its area of applicability.

```{r}
predict_optimism(d, folds, tgt, response = "z", coords = c("x", "y"), model = "idw")
```

## Recommend a validation scheme

`recommend_validation()` gives design-aware guidance; `rank_cv_schemes()` ranks
candidate schemes by how well they match deployment.

```{r}
recommend_validation(estimand = "prediction", design = "clustered", target = "grid")
rank_cv_schemes(d, tgt, response = "z", coords = c("x", "y"))
```

## Design where to validate

Given a budget of independent validation points, where should they go to estimate
true accuracy most precisely?

```{r}
design_validation(d, tgt, budget = 25, response = "z", coords = c("x", "y"))
```
