---
title: "Detecting and quantifying spatial leakage with spLeakage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Detecting and quantifying spatial leakage with spLeakage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4)
```

`spLeakage` **audits** a train/test split: it does not generate folds (that space is
covered by `CAST`, `blockCV`, `spatialsample`). Given a split, a prediction target,
and the data, it answers three questions:

1. **Is there leakage, and how much?** — the signed Spatial Leakage Index (`SLI`).
2. **So how inflated are my numbers?** — `estimate_optimism()`.
3. **What should I have done?** — `recommend_validation()`.

The guiding idea: *leakage and optimism are only well-posed relative to a declared
sampling design, estimand, and prediction target.*

```{r setup}
library(spLeakage)
```

## A clustered sample (the canonical leakage regime)

```{r data}
set.seed(3)
nc <- 8
centers <- cbind(runif(nc, 0.1, 0.9), runif(nc, 0.1, 0.9))
cl <- sample(nc, 500, replace = TRUE)
xy <- centers[cl, ] + cbind(rnorm(500, 0, 0.04), rnorm(500, 0, 0.04))
d <- data.frame(x = pmin(pmax(xy[, 1], 0), 1), y = pmin(pmax(xy[, 2], 0), 1))
d$z <- sin(2 * pi * d$x) + cos(2 * pi * d$y) + rnorm(500, 0, 0.15)
```

We will map wall-to-wall over the unit square:

```{r target}
s <- seq(0.025, 0.975, length.out = 20)
tgt <- prediction_target(grid = as.matrix(expand.grid(x = s, y = s)), type = "grid")
```

## 1. Detect leakage in a random 80/20-style 10-fold split

```{r detect}
folds <- sample(rep_len(1:10, nrow(d)))
lk <- detect_leakage(d, split = folds, target = tgt, response = "z",
                     coords = c("x", "y"), n_boot = 300)
lk
plot(lk)          # observed (CV) vs target (deployment) NN-distance ECDFs
```

A positive `SLI_rho` whose interval excludes zero means the split is optimistically
leaking. `plot(lk, which = "map")` shows *where*.

## 2. Quantify the optimism

```{r optimism}
opt <- estimate_optimism(d, split = folds, response = "z", coords = c("x", "y"),
                         control = "block")
opt
```

The reported (random-CV) error is compared against a leakage-controlled,
design-matched scheme. The gap is the optimism — here a large positive inflation.

## 3. Get a design-aware recommendation

Declare the estimand and design (the design *cannot* be inferred from coordinates):

```{r recommend}
recommend_validation(d, estimand = "prediction", design = "clustered",
                     target = "grid", coords = c("x", "y"))
```

For a **probability sample** the advice flips — random CV becomes correct and spatial
CV is over-pessimistic:

```{r recommend2}
recommend_validation(estimand = "prediction", design = "probability", target = "grid")
```

## Put it together: a scorecard

```{r report}
rec <- recommend_validation(d, estimand = "prediction", design = "clustered",
                            target = "grid", coords = c("x", "y"))
report_leakage(lk, optimism = opt, recommendation = rec)
```
