---
title: "Getting started: detecting and quantifying spatial leakage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started: detecting and quantifying spatial leakage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4)
```

## The problem in one paragraph

When you validate a spatial model with an ordinary **random** train/test split,
nearby points tend to land in *both* the training and the test set. Because spatial
data are autocorrelated (close locations have similar values), each test point
usually has a training neighbour just a short distance away. Your model is therefore
graded on a task close to **interpolation** — even though, at deployment, it must
predict at locations that are often *far* from any training data. The reported error
comes out too low: it is **optimistically biased**. This is *spatial information
leakage*, and it is why headline accuracy numbers for maps are routinely
over-confident.

## What `spLeakage` does (and does not) do

A mature ecosystem already *generates* leakage-resistant folds (`CAST`, `blockCV`,
`spatialsample`, `mlr3spatiotempcv`). `spLeakage` does something different: it
**audits a split you already have** (or a published result) and answers three
questions.

1. **Is there leakage, and how much?** → `detect_leakage()` returns the signed
   Spatial Leakage Index (SLI).
2. **So how inflated are my numbers?** → `estimate_optimism()` /
   `predict_optimism()` turn leakage into a percentage inflation of your accuracy.
3. **What should I have done instead?** → `recommend_validation()` gives
   design-aware advice (including *when spatial CV is the wrong choice*).

The single idea that ties these together:

> **Leakage and optimism are only well-posed relative to a declared sampling
> design, estimand, and prediction target.** The same split can be leaking,
> correct, or pessimistic depending on those three declarations.

So before measuring anything, you tell `spLeakage` *where the model will actually be
used* (the prediction target). Everything is measured against that.

```{r setup}
library(spLeakage)
```

## A worked example: a clustered survey

Real ecological, soil and disease-mapping samples are rarely a neat grid; they are
**clustered** — surveys concentrate where it is convenient or interesting to sample.
Clustering is the canonical leakage regime, so we simulate it: eight cluster centres
with points scattered tightly around each.

```{r data}
set.seed(3)
nc <- 8                                            # number of clusters
centers <- cbind(runif(nc, 0.1, 0.9), runif(nc, 0.1, 0.9))
cl <- sample(nc, 500, replace = TRUE)
xy <- centers[cl, ] + cbind(rnorm(500, 0, 0.04), rnorm(500, 0, 0.04))
d <- data.frame(x = pmin(pmax(xy[, 1], 0), 1), y = pmin(pmax(xy[, 2], 0), 1))
d$z <- sin(2 * pi * d$x) + cos(2 * pi * d$y) + rnorm(500, 0, 0.15)  # smooth spatial signal
```

The model is intended to produce a **wall-to-wall map** over the whole unit square,
not just to predict near the surveyed clusters. We declare that deployment target
explicitly — it is what the leakage will be measured against:

```{r target}
s <- seq(0.025, 0.975, length.out = 20)
tgt <- prediction_target(grid = as.matrix(expand.grid(x = s, y = s)), type = "grid")
tgt
```

`prediction_target()` builds the set of locations the model is deployed on. Use
`type = "grid"` for wall-to-wall mapping, `type = "newdata"` for a specific set of
prediction points, or `type = "interpolation"` for prediction *within* the sampled
footprint.

## Step 1 — Detect leakage

We grade a naive random 10-fold split against that target:

```{r detect}
folds <- sample(rep_len(1:10, nrow(d)))
lk <- detect_leakage(d, split = folds, target = tgt, response = "z",
                     coords = c("x", "y"), n_boot = 300)
lk
```

**How to read this output.**

* `SLI_rho (signed)` is the headline index. It is the difference in *retained
  spatial correlation* between cross-validation and deployment, on a
  $[-1, 1]$ scale:
  * **positive → optimistic leakage** (CV is easier than deployment; your error is
    too low),
  * **near zero → well matched** (CV imitates deployment),
  * **negative → pessimistic** (CV is *harder* than deployment; your error is too
    high — the Wadoux regime).
  The `90% CI` comes from the `n_boot` Monte-Carlo draws over variogram
  uncertainty. If the interval excludes zero, the leakage verdict is statistically
  resolved.
* `retained corr.` shows the two ingredients: `c_obs` is how much spatial
  correlation CV keeps to the training data, `c_pred` how much is available at
  deployment. `SLI_rho = c_obs - c_pred`. Here CV "sees" much more nearby signal
  than deployment will.
* `SLI_d (signed)` is the distance-form companion: `A` is literally how much
  *farther* deployment reaches than CV pretended (in coordinate units), normalised
  by the variogram range `phi`.
* `W (NNDM) / delta`: `W` is exactly the (unsigned) NNDM/kNNDM Wasserstein
  statistic, included for comparability. `delta` is the directionality ratio:
  near $+1$ means pure optimism, near $-1$ pure pessimism, near $0$ the ECDFs cross
  (mixed) — in which case you should look at the plot rather than trust a single
  number.

### See it, and see *where* it leaks

```{r plot-ecdf, fig.alt="Observed vs target nearest-neighbour distance ECDFs"}
plot(lk)                       # observed (CV) vs target (deployment) NN-distance ECDFs
```

The red curve (test→train distances during CV) sits to the **left** of the blue
curve (deployment→sample distances): CV test points are systematically closer to
their training data than deployment points will be. That gap *is* the leakage.

A per-fold breakdown, and a map of which points leak most, are one call away:

```{r summary}
summary(lk)                    # per-fold leakage contributions
```

```{r plot-map, fig.alt="Per-point leakage contribution in space"}
plot(lk, which = "map")        # red points contribute most leakage
```

This per-point/per-fold decomposition of an *arbitrary* split is something a
fold-generator cannot give you.

## Step 2 — Quantify the optimism

A leakage score is abstract; practitioners want the consequence in their own
currency. `estimate_optimism()` re-evaluates the model under a leakage-controlled,
design-matched scheme (here spatial block CV) and reports the gap.

```{r optimism}
opt <- estimate_optimism(d, split = folds, response = "z", coords = c("x", "y"),
                         control = "block")
opt
```

* `user CV error` is what your random split reported.
* `controlled error` is the (larger) error under the design-matched scheme.
* `optimism` is the difference, expressed as a percentage of the controlled error.
  A positive value means your reported accuracy is inflated by that much.

Optimism can legitimately be **negative** for a genuine probability sample — that is
*pessimism*, and the tool reports it as such rather than hiding it.

## Step 3 — Get a design-aware recommendation

What should you have done? The answer depends on facts the coordinates cannot
reveal — chiefly, the **sampling design**. A *designed* cluster sample and a
*convenience* cluster sample can have identical point patterns but require different
validation. So `spLeakage` asks you to **declare** the design; it never infers it.

```{r recommend}
recommend_validation(d, estimand = "prediction", design = "clustered",
                     target = "grid", coords = c("x", "y"))
```

Read the `recommended` schemes and the `avoid` line; the `clustering flag` is shown
as a **risk indicator only**, not as a basis for the recommendation.

Now watch the advice **flip** when the very same data are declared to be a
probability sample with a population-mean estimand — the case Wadoux et al. make,
where random validation is design-unbiased and spatial CV is over-pessimistic:

```{r recommend2}
recommend_validation(estimand = "population", design = "probability", target = "grid")
```

This is the reconciliation in action: neither "always use spatial CV" nor "never use
spatial CV" is correct — the right answer is a function of your declarations.

## Put it together: a submission-ready scorecard

`report_leakage()` assembles the diagnosis, the optimism, and the recommendation
into one object you can drop into a methods appendix.

```{r report}
rec <- recommend_validation(d, estimand = "prediction", design = "clustered",
                            target = "grid", coords = c("x", "y"))
report_leakage(lk, optimism = opt, recommendation = rec)
```

## Where to next

* **Leakage channels beyond distance** — grouped/duplicated locations, covariate
  space, time, and large-scale trend: `vignette("channels")`.
* **Quantify, correct, and design** — the fast emulator, the de-leaked estimate that
  corrects a published number, scheme ranking, and optimal validation-survey design:
  `vignette("quantify-correct")`.
```
